---
title: "Michael Jordan BioComp Assignment"
author: "Michael Jordan"
date: "2023-12-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction & Rationale

The malleefowl (Leipoa ocellata) is a large (~2kg) ground-dwelling bird endemic
to Australia.  It is a member of the Megapodiidae family, which incubate their
eggs in mounds that provide an external source of heat (rather than in nests
using body heat).  Within the Megapodiidae, the malleefowl is notable for being
the only species not to inhabit damp forests and for the size of their mounds.

Malleefowl mounds are large (~ 2 ft x 3 ft), scraped out of the ground by the
male, filled with leaf litter, and covered with an insulating layer of sand.
Solar radiation and fermentation of the organic material maintain a constant
temperature of 33°C within the nest, which is necessary for egg development.
Because sand and organic matter are required for mound construction, malleefowl
tend to inhabit arid to semi-arid shrublands and low woodlands.  Within these
habitats, malleefowl population concentration tends to be highest in areas with
higher rainfall and more fertile soil.

Although records suggests that the malleefowl was common prior to European
colonization, its population declined significantly post colonization due to
habitat loss and predation by introduced predators (e.g., red fox).  Today, the
malleefowl is listed as vulnerable on the IUCN Red List and under the Australian
Environment Protection and Biodiversity Conservation Act of 1999.

In 2007, the Australian government instituted a $2.4M AUD National Recovery Plan
for the malleefowl.  Although the report lacked sufficient evidence to
accurately assess malleefowl conservation status across Australia, it cited
steep declines in monitored populations -- as well as the fact that many known
populations were small and fragmented -- to justify action.  The NRP set
several objectives, including reducing habitat loss, reducing grazing pressure
by livestock, and reducing predation.

However, these initiatives are unlikely to be successful if climate change
renders current malleefowl habitat inviable.  Changes to rainfall and
temperature may be particularly significant, since they could impact breeding
success (e.g., if mound temperature falls below 33°C or the environment stops
producing sufficient leaf litter for mound construction).

This report aims to estimate the likely impact of climate change on malleefowl
species distribution in order to inform the malleefowl NRP.  If we expect
malleefowl habitat to expand or contract, NRP objectives may need to be
adjusted.


## Modelling Current and Future Malleefowl Species Distribution

All analysis was completed using R version 4.3.1.  Malleefowl occurrence data
was downloaded from the Global Biodiversity Information Facility (GBIF) database
using the Dismo R package.

```{r import}
# Load necessary libraries
library(dismo)
library(terra)
library(geodata)
library(sf)
library(rnaturalearth)
library(dplyr)
library(bestglm)

# Import data from GBIF and retain only relevant columns.
m.full <- dismo::gbif("Leipoa", "ocellata") # n = 10490
keep <- c("adm1", "adm2", "basisOfRecord", "cloc", "collectionCode",
          "coordinateUncertaintyInMeters", "day", "eventDate", "eventTime",
          "gbifID", "geodeticDatum", "georeferenceVerificationStatus",
          "higherGeography", "identifiedBy", "informationWithheld",
          "institutionCode", "lat", "locality", "lon", "month", "occurrenceID",
          "occurrenceRemarks", "protocol", "recordedBy", "year")
m <- m.full[, names(m.full) %in% keep]
```

Data was cleaned to include only trustworthy, geographically precise,
non-duplicative observations during the years covered by our historical climate
data (1970 - 2000) and plotted to verify a lack of outliers.

```{r clean}
# Clean data: 1) filter for observation occurrences, 2) remove occurrences
# without geo coords, 3) remove uncertain occurrences (unverified, obfuscated
# geo coords, > 1km geo coord uncertainty), 3) filter for records in baseline
# historical years.  Additionally, data were checked for incorrect lat/lon
#zeroing, but none was found.
m <- m[
  m$basisOfRecord %in% c("HUMAN_OBSERVATION", "MACHINE_OBSERVATION")
  & !is.na(m$lon)
  & !is.na(m$lat)
  & (
    is.na(m$georeferenceVerificationStatus)
    | m$georeferenceVerificationStatus == "verified")
  & is.na(m$informationWithheld)
  & (
    is.na(m$coordinateUncertaintyInMeters)
    | m$coordinateUncertaintyInMeters <= 1000)
  & !is.na(m$year)
  & m$year >= 1970
  & m$year <= 2000, ] # n = 774

# Drop observations that are duplicative at the day/lat/lon level.  Visual spot
# check confirmed identical metadata for duplicate events at this level.
m$dup <- duplicated(m[, c("eventDate", "lon", "lat")])
m <- m[m$dup == FALSE, ] # n = 731

# Declare geo coords extent and plot observations.
ext <- terra::ext(110, 155, -45, -8)
m.sf <- st_as_sf(m, coords = c("lon", "lat"), crs = 4326)
au <- ne_countries(country = "australia", scale = "large", returnclass = "sf")
plot(
  st_geometry(au),
  xlim = ext[1:2],
  ylim = ext[3:4],
  bg = "lightblue",
  col = "ivory",
  axes = TRUE)
plot(st_geometry(m.sf), add = TRUE, col = "black", border = NA)
```

To estimate current species distribution, historical climate data (1970 - 2000)
was pulled from WorldClim 2.1 using the geodata package.  Future climate data
was pulled for two time periods -- 2021 - 2040, 2041 - 2060 -- using the Hadley
GEM3 model and Shared Socioeconomic Pathway 2-4.5.  The two time periods were
chosen so that we could understand over what time scale we might see malleefowl
habitat shift.  SSP 245 was picked because it represents a "middle ground"
scenario and so should provide a conservative estimate of the extent of
malleefowl habitat changes.

Because malleefowl are likely to have elevation and soil habitat preferences,
the geodata package was used to pull elevation data from the Shuttle Radar
Topography Mission and % sand composition and carbon density (as a proxy for
soil richness) data from the SoilGRIDS database.

Climate and elevation data were pulled at a 2.5 arc-minute resolution (~1.3km at
Australian latitude).  Soil data were pulled at a 30 arc-second resolution and
aggregated to a 2.5 arc-minute resolution.

```{r gather}
# Pull historical and future climate data as well as elevation and soil data.
# Crop to Australia geo extent.
bioclim.h <- geodata::worldclim_global(var = "bio", res = 2.5, path = "data")
bioclim.h <- terra::crop(bioclim.h, ext)
bioclim.f1 <- geodata::cmip6_world(
  model = "HadGEM3-GC31-LL",
  ssp = "585",
  time = "2021-2040",
  var = "bioc",
  path = "data",
  res = 2.5)
bioclim.f1 <- terra::crop(bioclim.f1, ext)
bioclim.f2 <- geodata::cmip6_world(
  model = "HadGEM3-GC31-LL",
  ssp = "585",
  time = "2041-2060",
  var = "bioc",
  path = "data",
  res = 2.5)
bioclim.f2 <- terra::crop(bioclim.f2, ext)
elev <- geodata::elevation_global(res = 2.5, path = "data")
elev <- terra::crop(elev, ext)
sand <- geodata::soil_world(
  var = "sand",
  stat = "mean",
  depth = 60,
  path = "data")
sand <- terra::crop(sand, ext)
carbon <- geodata::soil_world(
  var = "ocd",
  stat = "mean",
  depth = 60,
  path = "data")
carbon <- terra::crop(carbon, ext)

# Aggregate soil rasters from 30 arc-second 2.5 arc-minute resolution.
sand <- terra::aggregate(sand, fact = 5, fun = mean)
carbon <- terra::aggregate(carbon, fact = 5, fun = mean)

# Align and simplify column names.
bioclim.names <- paste0("bio", 1:19)
names(bioclim.h) <- bioclim.names
names(bioclim.f1) <- bioclim.names
names(bioclim.f2) <- bioclim.names
names(elev) <- c("elev")
names(sand) <- c("sand")
names(carbon) <- c("ocd")
```

The Geoscience Australia Lambert projection was chosen due to [ADD RATIONALE]
and climate, elevation, and soil rasters were projected onto it and combined.

```{r project}
# Create a grid and project objects into the Geoscience Australia Lambert
#projection.
ntm.grid <- terra::rast(
  ext(-2740000, 2400000, -5252000, -944000), 
  res = 20000,
  crs = "EPSG:3112")
m.sf <- sf::st_transform(m.sf, crs = "EPSG:3112")
bioclim.h <- terra::project(bioclim.h, ntm.grid)
bioclim.f1 <- terra::project(bioclim.f1, ntm.grid)
bioclim.f2 <- terra::project(bioclim.f2, ntm.grid)
elev <- terra::project(elev, ntm.grid)
sand <- terra::project(sand, ntm.grid)
carbon <- terra::project(carbon, ntm.grid)

# Add elevation, sand, and carbon data to historical and future rasters.
bioclim.h <- c(bioclim.h, elev)
bioclim.h <- c(bioclim.h, sand)
bioclim.h <- c(bioclim.h, carbon)
bioclim.f1 <- c(bioclim.f1, elev)
bioclim.f1 <- c(bioclim.f1, sand)
bioclim.f1 <- c(bioclim.f1, carbon)
bioclim.f2 <- c(bioclim.f2, elev)
bioclim.f2 <- c(bioclim.f2, sand)
bioclim.f2 <- c(bioclim.f2, carbon)
```

A number of background psuedo-absence points equal to the observation count
(n = 731) were set using the Dismo package and plotted to visually inspect.

```{r absence}
# Create random psuedo absence geo coords.
n.pseudo <- nrow(m.sf)
land <- bioclim.h[["elev"]] > 0
pseudo.dismo <- dismo::randomPoints(
  mask = as(land, "Raster"),
  n = n.pseudo, 
  p = st_coordinates(m.sf))
pseudo.dismo <- sf::st_as_sf(
  data.frame(pseudo.dismo),
  coords = c("x", "y"),
  crs = 3112)

# Plot present and psuedo-absent points to visually inspect for errors.
# TODO: Add mask for non-AU
terra::plot(land, col = "grey", legend = FALSE)
terra::plot(st_geometry(m.sf), add = TRUE, col = "blue", cex = 1, pch = 2)
terra::plot(pseudo.dismo, add = TRUE, col = "black", cex = .1, pch = 19)
```

To reserve 20% of data for model evaluation, the Dismo package was used to add
quintiles and reserve one quintile for training.  Data were combined into a
single binomial presence/absence raster for data fitting.

```{r combine}
# Subdivide presence/absence for training/testing.
m.sf$kfold <- dismo::kfold(m.sf, k = 5)
pseudo.dismo$kfold <- dismo::kfold(pseudo.dismo, k = 5)

# Create presence/absence binomial dataset and add environmental data.
present <- terra::subset(m.sf, select = "kfold")
present$pa <- 1
absent <- pseudo.dismo
absent$pa <- 0
names(absent) <- c("geometry", "kfold", "pa")
sf::st_geometry(absent) <- "geometry"
pa.data <- rbind(present, absent)
envt.data <- terra::extract(bioclim.h, pa.data)
pa.data <- cbind(pa.data, envt.data)
```

The WorldClim data provides 19 climate measurements, many of which are
conceptually similar (e.g., Precipitation of Wettest Quarter vs Precipitation of
Wettest Month).  The following variables were deemed worth of consideration to
include in a model since they could plausibly impact the malleefowl's lifecycle:
Annual Mean Temperature, Isothermality, Max Temperature of Warmest Month, Min
Temperature of Coldest Month, Annual Precipitation, Precipitation of Wettest Month, Precipitation of Driest Month (Bio1, Bio3, Bio5, Bio6, Bio12, Bio13, Bio14).

Including elevation, sand composition of soil, and carbon density of soil, we
considered 10 potential coefficients to model.

```{r column_filter}
# Select relevant columns from historical and future rasters
bioclim.h <- terra::subset(
  bioclim.h,
  c("bio1", "bio3", "bio5", "bio6", "bio12", "bio13", "bio14", "elev", "sand",
    "ocd"))
bioclim.f1 <- terra::subset(
  bioclim.f1,
  c("bio1", "bio3", "bio5", "bio6", "bio12", "bio13", "bio14", "elev", "sand",
    "ocd"))
bioclim.f2 <- terra::subset(
  bioclim.f2,
  c("bio1", "bio3", "bio5", "bio6", "bio12", "bio13", "bio14", "elev", "sand",
    "ocd"))
pa.data <- terra::subset(
  pa.data,
  select = c("kfold", "pa", "bio1", "bio3", "bio5", "bio6", "bio12", "bio13",
             "bio14", "elev", "sand", "ocd"))
```

The bestglm package was used to assess which coefficients to include in a
binomial GLM using Aikake Information Criteria.

```{r aic}
# Subset presence/absence data and run bestglm to select best model per AIC.
fit <- as.data.frame(pa.data)
fit <- fit[, !(names(fit) %in% c("geometry", "kfold"))]
fit <- relocate(fit, pa, .after = last_col())
bestglm(fit, IC = "AIC", family = binomial)
```

Per AIC, the best fitting model included 9 of the variables (dropping Bio3,
Isothermality).

```{r fitting}
# Fit the GLM.
glm.model <- glm(
  pa ~ bio1 + bio5 + bio6 + bio12 + bio13 + bio14 + elev + sand + ocd,
  data = pa.data, 
  family = binomial(link = "logit"),
  subset = kfold != 1)

# Summarize and plots model results
summary(glm.model)
par(mar = c(3, 3, 1, 1), mgp = c(2, 1, 0))
dismo::response(
  glm.model,
  fun = function(x, y, ...) predict(x, y, type = "response", ...))
```

The model results demonstrate that all coefficients were statistically
significant, except for Bio6 (Min Temperature Of The Coldest Month).  Examining
the response graphs, the direction of relationships makes biological sense
(e.g., presence increases with sand, rainfall, decreases with elevation).

Finally, the model was fit on historical data and ROC curves plotted.

```{r roc}
# Create a prediction layer and evaluate the results.
glm.pred <- terra::predict(bioclim.h, glm.model, type = "response")
test.present <- sf::st_coordinates(subset(pa.data, pa == 1 & kfold == 1))
test.absent <- sf::st_coordinates(subset(pa.data, pa == 0 & kfold == 1))
glm.eval <- dismo::evaluate(
  p = test.present,
  a = test.absent,
  model = glm.model,
  x = bioclim.h)
print(glm.eval)

# Plot ROC curves and kappa.
max.kappa <- plogis(threshold(glm.eval, stat = "kappa"))
par(mfrow = c(1, 2))
plot(glm.eval, "ROC", type = "l")
plot(glm.eval, "kappa", type = "l")
abline(v = max.kappa, lty = 2, col = "blue")
```

The area under the ROC curve was .977, indicating little trade-off between
sensitivity and specificity across threshold of presence as determined by the
model.

Finally, estimated current distributed was plotted along with future
distribution for 2021 - 2040 and 2041 - 2060.

```{r roc}
# Plot species distribution maps
# TODO: Mask non-australia islands 
par(mfrow = c(1, 3))
glm.map <- glm.pred >= max.kappa
plot(glm.map, legend = FALSE, col = c("grey", "red"))

glm.pred.future1 <- predict(bioclim.f1, glm.model, type = "response")
glm.map.future1 <- glm.pred.future1 >= max.kappa
plot(glm.map.future1, legend = FALSE, col = c("grey", "red"))

glm.pred.future2 <- predict(bioclim.f2, glm.model, type = "response")
glm.map.future2 <- glm.pred.future2 >= max.kappa
plot(glm.map.future2, legend = FALSE, col = c("grey", "red"))
```

## Statistical Analysis of Malleefowl Encroachment On Cropland



## Summary

